---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installing and using the package

To install the package, you can use the `remotes` package:

```{r, eval = FALSE}
remotes::install_github("ndpvh/denoiser")
```

To load the package, use `library` 

```{r}
library(denoiser)
```


## Overview of the package

The `denoiser` package was introduced to meet the need for a systematic way to add realistic noise to simulated positional data and, similarly, to filter out this noise within the same type of data. This need arose within our own research group with the definition of recovery studies for the [Minds for Mobile Agents](https://www.ampl-psych.com/projects/minds-for-mobile-agents/) pedestrian model with data that were simulated through the [predped](https://github.com/ndpvh/predped) package, but the package may apply more broadly to other research goals. Note, however, that the current implementation relies heavily on a set of calibration studies that we have carried out in our own lab, the results of which can be found in [this repository](https://github.com/ndpvh/calibration-movement-data).

The package revolves around two primary functions. The `noiser` function adds noise to data that are provided by the user. Within the package, we refer to function that adds a particular type of noise as the  "measurement model", reflecting that we define how positional measurements relate to actual positions. Natively, we provide two such measurement models: One that assumes that measurement error is independent over time and one that does assume a temporal component to the measurement error, both of which are defined more extensively in the vignette on [Adding noise](https://ndpvh.github.io/denoiser/articles/noiser_vignette.html). However, we also allow users to specify the measurement model themselves, making the package more broadly applicable. 

The `denoiser` function filters out noise from the data that are provided by the user. By default, filtering will be through a Kalman filter that makes use of what we internally refer to as the "Kalman model". Currently, we only provide one such model, namely the constant velocity model, the specification of which you can find in the vignette on [Filtering data](https://ndpvh.github.io/denoiser/articles/denoiser_vignette.html). Users can specify their own model, however, again making the package more broadly applicable. Besides the Kalman filter, the package also natively supports binning one's data, although it doesn't do so by default. Finally, users can provide their own filters to the package, further adding to the package's usefulness.
